---
title: "Interactive mapping with RShiny"
subtitle: "*Engaging end-users with research*"
author: "Christina Buelow"
institute: "Made for RLadies Brisbane (May 2022)"
#date: "2022/05/26 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: left, top

# A journey in maps and apps...

1. The basics of programming for interactivity: *Reactivity*

--

2. R package options for mapping interactively in R, with pros and cons
--

3. Make an interactive map and share in a web app
--

.footnote[[*] This journey is thanks to the awesome book [Mastering Shiny](https://mastering-shiny.org/index.html)]

---
# Why interactive and why Shiny?

--
#### Open Science
  - Interactive web apps are a great way to make our research **open**, **transparent**, and **engaging**

--

  - Move beyond static papers

--

#### Shiny 
  
  - Performative apps without HTML, javascript, CSS...

--

  - Range of users - academics to tech companies 
  
--
  
  - BUT, needs to be hosted on a server instead of directly on a web browser

--
#### Server options:
  - shinyapps.io: free or subscription options
  - [See here for more info on Rshiny servers](https://shiny.rstudio.com/articles/shiny-server.html)
  - There are other 'non-R' server options out there too
---
# App fundamentals

Interactivity requires *Reactivity*

What is **reactive programming**?

--

  - Outputs *react* to inputs

--
    - Means computations are only done when necessary (moving beyond **functions** to **reactions**)

--
    - Reactive expressions are **lazy** and **cached**

--

  - It's different from how we normally code our scripts:
  
    - Commands vs. recipes
    - Imperative vs. declarative
    - Assertive vs. passive-aggressive

???
In imperative programming, you issue a specific command and it’s carried out immediately. This is the style of programming you’re used to in your analysis scripts: you command R to load your data, transform it, visualise it, and save the results to disk.

In declarative programming, you express higher-level goals or describe important constraints, and rely on someone else to decide how and/or when to translate that into action. This is the style of programming you use in Shiny.

With imperative code you say “Make me a sandwich”8. With declarative code you say “Ensure there is a sandwich in the refrigerator whenever I look inside of it”. Imperative code is assertive; declarative code is passive-aggressive.

lazy and cached: minimal set of computations to update ouputs based on inputs
---

# Shiny

{shiny} provides a framework for **reactive programming**:
  - User-interface (ui): how it looks. Takes *inputs* and displays *outputs*
  - Server: has the recipe to turn *inputs* into *outputs*

--

*hint*: if you have {shiny} installed, just type `shinyapp` in your R script to insert boilerplate shiny app code and get started quickly

```{r}
library(shiny)

ui <- fluidPage(
  
)

server <- function(input, output, session) {
  
}

shinyApp(ui, server)
```

???

* Here pause to do this together in R. Show what fluidPage means using shiny cheat-sheet.
* ui and server can be stored as different objects in an app.R script and deployed with shinyApp function
* or, for more complex apps, split and save as separate ui.R and server.R scripts
* Discuss input (widget) vs. output (render) functions in R
* Run the example in the cheet sheet, just to show how a basic app works
---

# A simple histogram

```{r}
hist(rnorm(25))
```
---

# A reactive histogram

```{r}
library(shiny)

ui <- fluidPage(
  numericInput(inputId = "n", "Sample size", value = 25), 
  plotOutput(outputId = "hist")
)

server <- function(input, output, session) {
  output$hist <- renderPlot({
    hist(rnorm(input$n))
    })
}

shinyApp(ui, server)
```
--
.footnote[[ ] We won't cover an additional reactive tool, *reactive expressions*, but find more information [here](https://mastering-shiny.org/reactivity-intro.html). *Reactive expressions* can be used to eliminate redundant code in your app, thereby improving efficiency. ]

???
One more important thing about input: it’s selective about who is allowed to read it. To read from an input, you must be in a reactive context created by a function like renderText() or reactive().

There’s one more important component that you’ll see in the reactive graph: the reactive expression. We’ll come back to reactive expressions in detail very shortly; for now think of them as a tool that reduces duplication in your reactive code by introducing additional nodes into the reactive graph.
---

# But what about maps?

My go-to for mapping in R these days is {tmap}

One of the reasons is that I can make either interactive or static, publication-quality maps with ease. Just set the 'mode'.

```{r, warning = F}
library(tmap)
data('World')

tmap_mode('view')
qtm(World, 'name') # qtm stands for 'quick thematic mapper'
```
---
# Share with shiny

If we're in a hurry, we can put our interactive {tmap} in a {shiny} app to share

```{r}
library(shiny)
library(tmap)
data('World')

ui <- fluidPage(
  tmapOutput('map')
)

server <- function(input, output, session) {
  output$map <- renderTmap({
    qtm(World, 'name')})
}

shinyApp(ui, server)
```

---
# Add some more *reactivity*

Let's add a widget to allow users to choose their country from a list

```{r}
library(shiny)
library(tmap)
library(dplyr)
data('World')

ui <- fluidPage(
  tmapOutput('map'),
  selectInput('var', 'Select country',
              choices = c('Global', as.character(World$name)),
              selected = 'Global')
)

server <- function(input, output, session) {
  output$map <- renderTmap({
    if(input$var == 'Global'){
      qtm(World, 'name')
    }else{
      qtm(filter(World, name == input$var))
    }
  })
}

shinyApp(ui, server)

```

---

# Customising and scaling your interactive map

If you want to **customise** your map with more *features*, the [leaflet](https://rstudio.github.io/leaflet/) R package offers some nice functionality (there are others too)

--

{tmap} and {leaflet} both use **'scalable vector graphics'** (SVG) to visualise maps on the web

--

  - This is fine for simple apps, but if you need to render large datasets or do complex computations they will be really slow

--

**Scaling-up** your map app will require graphics to be rendered in **'WebGL'** instead of SVG

--

  - There are several R packages to try including [mapdeck](https://symbolixau.github.io/mapdeck/articles/mapdeck.html) and [rdeck](https://github.com/anthonynorth/rdeck)
--

  - These usually require API tokens to access basemaps from [mapbox](https://www.mapbox.com/)

--

Should I go **SVG** or **WebGL**?

  - Generally I've found there is a **feature vs. speed trade-off**
  - Recommend start with **SVG** and if too slow then move to **WebGL** 

???

What are my interactive web mapping options?

Interactive web visualisations:
Options for graphics on the web:

SVG (Scalable vector graphics) - simplest option, but not scalable, can’t handle large datasets. This is leaflet
WebGL - good if you have lots of data to render in browser (points/polygons), need to perform lots of data calculations, want to animate

WebGL can be used to visualize large, complex graphics with excellent performance. WebGL provides an API that allows you to write low-level code for creating rasterized images. This code is run on your computer's GPU instead of CPU, allowing large data to be processed in parallel.

WebGL = Performance!!

But complex b/c works with low level code

So pick what you need depending on use-case. Only go up the ladder when not performing.

—
Mapbox - base map provider (also Google Maps, ArcGIS, etc)


Leaflet vs leafgl vs mapveiew vs map deck vs rdeck

Map deck
- No feature click popsps
- Needs an API key
- Offers some cool functions
    - Heat map, grid map column plots
- Can render large datasets

Mapdeck has more fine grained controlled than map view

Rdeck

In short it's a similar premise to {mapdeck} but implemented in a different way which comes with many benefits. It supports much more of http://deck.gl (almost all of it), it uses less data, has a ggppot2 scale-style approach to legends, built in layer selector...

Mapbox - base maps

API - Application programming interface

---

# Leaflet example



---

# Tips and tricks

Lots of complex polygons?
Use [rmapshaper](https://cran.r-project.org/web/packages/rmapshaper/vignettes/rmapshaper.html) to simplify, this can really **speed up** a slow map app

Have you reached the **'cliff of complexity'**?
Check out [Engineering production grade shiny apps](https://engineering-shiny.org/successful-shiny-app.html)


???
- Modules
- Caching
-etc....
To reiterate the section on “Improving performance and scalability” in shiny from Cheng (2018a), you have a number of tools available to address performance:

The profvis package for profiling code.
Cache computations ahead-of-time.
Cache computations at run time.
Cache computations through chaining reactive expressions.
Leverage multiple R processes and/or servers.
Async programming with promises


---
Let's get started

Open 'app.R' file
File | New Project, then selecting New Directory and Shiny Web Application.

Basics 
shiny R package
Define ui - (i.e. the HTML webpage human interact with)
Define server - what can our app do?
Execute the app (shinyApp function)
---
# Setting up your UI

Layout function fluidPage - layout function, what goes where?

How do we let end-users interact? Inputs.
Widgets - e.g., selectInput, etc.

How does shiny react to end-user inputs?
Outputs. We define those in the UI too.

(Fancy ways to generate HTML)
---
# Bringing outputs to life

Reactive programming = interactive

`It’s like the difference between giving someone a recipe versus demanding that they go make you a sandwich.`

Give a recipe

When inputs change, app will react and outputs will change based on recipe.

Server is essentially a function with input, output, session arguments.

Server connects the inputs and outputs - fully functional app.

Provide a recipe for each shiny output with a specific ID.

Render functions wrap up the recipe.

---

# Duplicated code

Usually capture value using a variable, or capture computation with a function.

Shiny answer: reactive expressions

Assign reactive expression to a variable, and can call it like a function

